GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2022-12-13 writeUps/paper.tex

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{numdef}

@@

Randomized experiments easily admit subgroup analysis---estimation of treatment effects for a subset of the study population---when those subgroups are defined at baseline, and not themselves a function of treatment assignment.
A simple, popular approach to estimating subgroup effects is to use ordinary least squares (OLS) regression, regressing the outcome on a treatment indicator interacted with an indicator for subgroup membership.
However, when subgroup membership is a function of a post-treatment variable, then it may itself be affected by the treatment.
Contrasts between treatment and control conditions within those subgroups may not have a causal interpretation at all---and when they do, the same techniques that work for subgroups defined at baseline (such as OLS with interactions) will often produce inconsistent estimates.
Principal stratification \citep{frangakis} is a broad framework for defining subgroup effects for post-treatment subgroups, based on the notion of \emph{potential} subgroups called principal strata, that is, groups of subjects defined based on the values a post-treatment variable would potentially take if the subject is assigned to one treatment condition or another.
Since principal strata membership is based on subjects' responses to both factual and counter-factual treatment assignments, it is, in general unobserved.
Nevertheless, researchers have devised estimators of principal effects, as average effects within principal strata are called, to address a broad range of applied statistical problems, including non-compliance in randomized experiments, the evaluation of surrogate outcomes \citep{li2010bayesian}, the investigation of causal mechanisms \citep{lidsayPage}, and sample attrition \citep{zhangRubin, ding2011}.

@@

In a separate, if related, line of research, \citet{jo2002}, \citet{ding2011}, \citet{jiangDing2021}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.
This paper expands on those identification results by presenting a new set of principal effect estimators, and associated standard error estimates, based on a generalized estimating equations (or M-estimation or method-of-moments) \citep{stefanskiBoos}.
It turns out that, under certain general conditions, principal effects can be estimated with a simple, two-step procedure: first, estimate principal scores with logistic regression or another M-estimator; then, impute unknown principal subgroup membership indicators with principal scores and estimate principal effects using OLS with interactions, just as in the pre-treatment subgroup case.
We will refer to the estimators presented here as ``GEEPERs'': generalized estimating equations for principal effects using regressions.
GEEPERs estimates rely on correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the distribution of regression errors.
Unlike principal score weighting, GEEPERs does not assume principal ignorability, and admits to sandwich-style standard error estimates, which we provide in the appendix.

@@

\section{Background}
In a randomized experiment with two conditions, let $Z_i=0$ if subject $i$, $i=1,\dots,n$ is assigned to the treatment condition, and let $Z_i=0$ otherwise.
If interest is in the effect of $Z$ on an outcome $Y$, under the ``stable unit treatment value assumption'' \citep{air} of no interference between units and no hidden versions of the treatment, define potential outcomes \citep{splawa1990application,air} $\yti$ as the value of $Y$ that $i$ would exhibit if $Z_i=1$, and let $\yci$ be the value if $Z_i=0$. Then the observed value of $Y$ satisfies $Y_i=Z_i\yti+(1-Z_i)\yci$, and the treatment effect of $Z$ on $Y$ for subject $i$ may be defined as $\tau_i=\yti-\yci$. The average treatment effect (ATE) is $\EE[\tau]$. %

@@

\subsection{Estimating Principal Effects}
In estimating principal effects for a binary $S$, it will be useful to define four conditional means:
\begin{equation}\label{eq:mus}
  \begin{split}
    \muc0&=\EE[Y_C|\st=0]\\
    \muc1&=\EE[Y_C|\st=1]\\
    \mut0&=\EE[Y_T|\st=0]\\
    \mut1&=\EE[Y_T|\st=1]
  \end{split}
\end{equation}
Then, since
\begin{equation*}
  \begin{split}
    \eff0&=\mut0-\muc0\\
    \eff1&=\mut1-\muc1
  \end{split}
\end{equation*}
estimating principal effects is equivalent to estimating the four conditional means.

@@

Under Assumptions \ref{ass:sm} and \ref{ass:rand}, conditional means $\mut0$ and $\mut1$ are non-parametrically identified.
In particular, the means of observed outcomes for subjects with $Z=1$ and $S=0$ or $S=1$ are unbiased for $\mut0$ and $\mut1$, respectively, since when $Z_i=1$, $Y_i=\yti$ and $S_i=\sti$.
In contrast, $\muc0$ and $\muc1$ are not fully identified without further assumptions because $\sti$ and $\yci$ are never observed simultaneously (though they may be nonparametrically bounded; see, e.g., \citealt{bounding}).
When $S$ is a measure of compliance, and $\tau_i=0$ whenever $\sti=0$ (the ``exclusion restriction''), $\muc1$---and hence $\eff1$---may be estimated with instrumental variables methods \citep{air}.
In this case, $\eff1$ is often called the ``local average treatment effect'' or the average effect of the treatment on the treated.
We will briefly review two approaches to estimating $\muc0$ and $\muc1$ that do not assume the exclusion restriction---finite mixture modeling and weighting---with some comments on the role for covariates in-between.

\subsubsection{Finite Mixture Modeling}
The finite mixture modeling approch to estimating $\muc0$ and $\muc1$ \citep{imbens1997bayesian} assumes that, conditional on $\st$, $Y_C$ is normally distributed.
Then the probability density of $Y_C$ may be written as
\begin{equation}\label{eq:mixtureModel}
  f_{Y_C}(y)=Pr(\st=0)\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)+Pr(\st=1)\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
\end{equation}
where $\phi(\cdot)$ is the standard normal density function, and $\sigma_C^0>0$ and $\sigma_C^1>0$ are standard deviations, which are sometimes assumed equal.
The probabilities $Pr(\st=0)$ and $Pr(\st=1)$ may be estimated first using data from the treatment group, since Assumption \ref{ass:rand} implies that $Pr(\st=1|Z=1)=Pr(\st=1)$.
Given these estimates and \eqref{eq:mixtureModel}, parameters $\muc0$ and $\muc1$ may be estimated using maximum likelihood, method of moments, or Bayesian techniques.

@@

\subsection{Principal Scores}
The principal score \citep[e.g.][]{jo2009use} is defined as
\begin{equation}\label{eq:pscore}
  \ppi=Pr(\sti=1|\bxsi)
\end{equation}
where $\bxsi\subseteq\bm{x}_i$, a not-necessarily-proper subset of covariates used to model $S$.
Note that under randomization, $Pr(\sti=1|\bxsi,Z_i)=Pr(\sti=1|\bxsi)$, so a model for principal scores can be estimated using data from the treatment group and extrapolated to the control group.
Assume a model for principal scores:
\begin{equation}\label{eq:prinScoreMod}
  \ppip{\bm{\alpha}}=f(\bxs;\bm{\alpha})
\end{equation}
with parameter vector $\bm{\alpha}$. An analyst may estimate $\bm{\alpha}$ by fitting model \eqref{eq:prinScoreMod} using observed $\bm{S}$ and $\bxs$ for subjects with $Z=1$, and then compute estimated principal scores $\ppip{\bm{\hat{\alpha}}}$ for subjects with $Z_i=1$.

@@

In order for principal scores to be useful, they need to vary with $\bm{x}$. Otherwise, $\bm{x}$ is uninformative about $\st$. In particular, we will assume the following:
\begin{ass}[Variable Principal Scores]\label{ass:vps}
 $\pp$ takes at least 3 distinct values
\end{ass}
In some estimators we will discuss below, two distinct values suffice. For a broader discussion, see \citet{ding2011,jiangDing2021}.

Principal scores can potentially improve inference in a finite mixture model, by re-writing \eqref{eq:mixtureModel} as
\begin{equation}\label{eq:mixtureModelPS}
  f_{Y_C}(y)=(1-\ppip{\bm{\alpha}})\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)+\ppip{\bm{\alpha}}\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
\end{equation}
Ideally, a statistician will estimate parameters $\muc0$, $\muc1$, $\mut0$, $\mut1$, $\sigma^0_C$, $\sigma^1_C$, and $\bm{\alpha}$ simultaneously, using maximum likelihood or Bayesian techniques.

Alternatively, assume ``principal ignorability'' \citep{jo2009use,dingLu}:
\begin{ass}[Principal Ignorability]\label{ass:PI}
\begin{equation}\label{eq:pi}
  Y_{C}\independent \st |\bxs
\end{equation}
\end{ass}
or a somewhat weaker version, $\EE[Y_C|\st,\bxs]=\EE[Y_C|\bxs]$ \citep{feller2017principal}.
That is, the principal stratum is unrelated to control potential outcomes conditional on covariates.
Principal ignorability \eqref{eq:pi} is reminiscent of ignorability assumptions typical in observational studies \citep[e.g.][]{rosenbaum2010design}.
Both require a rich set of covariates and background knowledge of the determinants of $S$; in particular, an unobserved covariate $u$ that is correlated with both $\st$ and $Y_C$ would invalidate \eqref{eq:pi}.

@@

We begin by introducing an assumption that we term ``Covariate Ignorability,'' or CI. It is a slighty weaker version of \citet{jiangDing2021}'s ``auxiliary independence'' assumption.
\begin{ass}[Covariate Ignorability]\label{ass:ci}
\begin{equation}\label{eq:assumption}
\EE[\yci|\bxsi,\sti]=\EE[\yci|\sti]=\muc0\text{ or }\muc1
\end{equation}
\end{ass}
i.e. $Y_C$ is mean-independent of $\bxs$ conditional on $\st$.
Under CI, covariates are not informative of the mean of $Y_C$ within principal strata.
As stated above, CI will rarely be plausible.
However, in some circumstances researchers can identify a subset of observed covariates that satisfy CI, and use those covariates in estimation.

It turns out that under strong monotonicity, randomization, and CI, and given a set of principal scores, the two principal effects $\eff0$ and $\eff1$ can be estimated by a simple OLS regression.
To see how, we'll %first
derive the estimating equations for $\muc0$, $\muc1$, $\mut0$, and $\mut1$, and show that they are equivalent to estimating equations for a model fit by OLS.% assuming that principal scores are known exactly, and then generalize to the case in which they are estimated using data from the treatment group.
The details of the calculations and the proofs are in the appendix.

The argument stems from a set of expressions for expectations, summarized in the following lemma:
\begin{lemma}\label{lemma:expectation}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, and \ref{ass:ci}, %if principal scores $\ppi$, for $i:Z_i=0$ are known exactly, then:
  \begin{equation*}
    \begin{split}
      \EE[Y|Z=0]=&\muc0+\EE\pp(\muc1-\muc0)\\
      \EE[Y\pp|Z=0]&=\EE[\pp]\muc0+\EE[\pp^2](\muc1-\muc0)\\
      \EE[Y|Z=1]&=\mut0+\EE[\st](\mut1-\mut0)\\
      \EE[YS|Z=1]&=\EE[\st]\mut0+\EE[\st^2](\mut1-\mut0)
    \end{split}
  \end{equation*}
\end{lemma}
Since $\st=\st^2$, the expression for $\EE[\st Y_T]$ in Lemma \ref{lemma:expectation} can be simplified to $\EE[\st]\mut1$; however, we retain the unsimplified version, since its similarity with the analogous expression when $Z_i=0$ simplifies the reasoning below.

@@

\begin{prop}\label{prop:reg1}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, \ref{ass:ci}, and suitible regularity conditions \citep[e.g.][p. 327]{boosStefanskiBook}, if %principal scores are known exactly,
  $\ri$ is defined as in \eqref{eq:ri} and the model
\begin{equation}\label{eq:regression0}
  \EE[Y_i|Z_i,\ri]=\beta_0+\beta_1\ri+\beta_2Z_i+\beta_3Z_i\ri
\end{equation}
is fit with OLS, yielding coefficient estimates $\hat{\beta}_k$, $k=0,\dots,3$, then
\begin{equation}
  \begin{split}
    \hat{\eff0}&\equiv \hat{\beta}_2\\
    \hat{\eff1}&\equiv \hat{\beta}_3-\hat{\beta}_2
  \end{split}
\end{equation}
are M-estimators, with $\hat{\eff0}\rightarrow_p\eff0$ and $\hat{\eff1}\rightarrow_p\eff1$ as $n\rightarrow\infty$
\end{prop}

@@

If principal scores were known exactly, the M-estimation structure of $\hat{\eff0}$ and $\hat{\eff1}$ implies that their sampling variance could be estimated by the standard sandwich estimator for OLS \citep{stefanskiBoos}.
This will rarely be the case.
However, if the principal scores are modeled as \eqref{eq:prinScoreMod}, and parameters $\bm{\alpha}$ are estimated by M-estimation, with estimating equations $\sum_{i: Z_i=1}\Omega_i=\bm{0}$, then the estimating equations for the principal score model and principal effect estimation can be stacked \citep[c.f.][]{boosStefanskiBook}, as
\begin{equation}\label{eq:stacked}
  \sum_{i=1}^n \begin{pmatrix}
    Z_i\Omega_i(\bxsi,S_i;\bm{\alpha})\\
    \Psi_i\left(Y_i,Z_i,\ri(\bm{\alpha}); \bm{\beta}\right)\end{pmatrix}=\begin{pmatrix} \bm{0}\\\bm{0}\end{pmatrix}
\end{equation}
where $\ri\left(\bm{\alpha}\right)=S_iZ_i+\ppip{\bm{\alpha}}(1-Z_i)$ emphasizes $\ppi$'s dependence on $\bm{\alpha}$.
In practice, the principal score model \eqref{eq:prinScoreMod} could be fit first, yielding estimates $\bm{\hat{\alpha}}$, and model \eqref{eq:regression0} could be fit next, using $\ri=\ri\left(\bm{\hat{\alpha}}\right)$.
Together, the vector of estimates $[\bm{\hat{\alpha}},\bm{\hat{\beta}}]$ represents a zero of the stacked estimating equations \eqref{eq:stacked}.

Then, the standard error matrix for $[\bm{\hat{\alpha}},\bm{\hat{\beta}}$ can be estimated as \citep[][ch. 7]{boosStefanskiBook}:
\begin{equation}\label{eq:sandwich}
  \widehat{var}(\bm{\hat{\beta}})=A^{-1}BA'
\end{equation}
Where
\begin{equation*}
  A=\sum_i\frac{\partial}{\partial [\alpha,\beta]'} [Z_i\Omega_i,\Psi_i]'
\end{equation*}
and
\begin{equation*}%\label{eq:Bmat}
  B=\sum_i [Z_i\Omega_i, \Psi_i]'[Z_i\Omega_i, \Psi_i]
\end{equation*}
where $[Z_i \Omega_i, \Psi_i]$ is the stacked vector of estimating equations evaluated at $Z_i$, $\bxsi$, $S_i$, and $Y_i$.
The appendix gives details of this calculation, giving a consistent variance estimator.

To summarize, under strong monotonicity, randomization, and covariate ignorability \eqref{eq:assumption}, principal effects can be estimated with the following procedure:
\begin{enumerate}
\item Estimate parameters $\bm{\alpha}$ as $\bm{\hat{\alpha}}$ by fitting model \eqref{eq:prinScoreMod} using data from subjects with $Z=1$.
\item Estimate principal scores for subjects with $Z_i=0$ as $\ppip{\bm{\hat{\alpha}}}$
\item Set $\ri=Z_iS_i+(1-Z_i)\ppip{\bm{\hat{\alpha}}}$
\item Using OLS, fit model \eqref{eq:regression0}
\item Estimate $\hat{\tau^0}=\hat{\beta}_2$ and $\hat{\tau^1}=\hat{\beta}_3-\hat{\beta}_2$
\end{enumerate}

\subsection{Relaxing Covariate Ignorability with Outcome Modeling}\label{sec:regression}
In most applications, identifying a set of covariates $\bxs$ satisfying CI will be difficult or impossible.
Fortunately, CI may be relaxed or obviated by regression (see \citealt[][\S 3.4]{jiangDing2021} for an analogous identification result).
Let $\bxy\in \mathbb{R}^k$, be a vector of covariates (possibly the same as $\bxs$); $\bxy$ can contain transformations and interactions of baseline covariates, allowing for a wide range of models.
Then note that $\st$ is, essentially, a covariate, so by randomization, (Assumption \ref{ass:rand}), $\EE[\bxy|\st]=\EE[\bxy|\st,Z=0]=\EE[\bxy|\st,Z=1]$.
Then information about the stratum-specific means of $\bxy$ can be extrapolated from the treatment to the control group.
This leads to the following proposition:
\begin{prop}\label{prop:interactions}
  Say, for $i=1,\dots,n$, principal scores $\ppi$ are generated as \eqref{eq:pscore}, with parameters $\bm{\alpha}$ identified and estimable with M-estimation, and outcomes are generated as
  \begin{equation}\label{eq:interaction}
    Y_i=\beta_0+\beta_1\sti+\beta_2 Z_i+\beta_3Z_i\sti+\bm{\gamma_1}'\bxy_i+\bm{\gamma_2}'\bxy_i\sti+
    \bm{\gamma_3}'\bxy Z_i+\bm{\gamma_4}'\bxy_i Z_i\sti+\epsilon_i
  \end{equation}
  with $\EE[\epsilon_i]=0$, $\{Z_i,\sti,\bxy_i\}\independent \epsilon_i$, and $\epsilon_i\independent\epsilon_j$ when $i\ne j$.

>>>>

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{numdef}

@@

Randomized experiments easily admit subgroup analysis---estimation of treatment effects for a subset of the study population---when those subgroups are defined at baseline, and not themselves a function of treatment assignment.
%A simple, popular approach to estimating subgroup effects is to use ordinary least squares (OLS) regression, regressing the outcome on a treatment indicator interacted with an indicator for subgroup membership.
However, when subgroup membership is a function of a post-treatment variable, then it may itself be affected by the treatment.
Contrasts between treatment and control conditions within those subgroups may not have a causal interpretation at all---and when they do, the same techniques that work for subgroups defined at baseline %(such as OLS with interactions) 
will often produce inconsistent estimates.

Principal stratification \citep{frangakis} is a broad framework for defining subgroup effects for post-treatment subgroups. It relies on the notion of \emph{potential} subgroups called principal strata: groups of subjects defined based on the values a post-treatment variable would potentially take if the subject were assigned to one treatment condition or another.
Since principal strata membership is based on subjects' responses to both factual and counter-factual treatment assignments, it is in general unobserved.
Nonetheless, researchers have devised estimators of principal effects, as average effects within principal strata are called, to address a broad range of applied statistical problems, including non-compliance in randomized experiments \citep{air}, the evaluation of surrogate outcomes \citep{li2010bayesian}, the investigation of causal mechanisms \citep{lidsayPage}, and sample attrition \citep{zhangRubin, ding2011}.

@@

In a separate, if related, line of research, \citet{jo2002}, \citet{ding2011}, \citet{jiangDing2021}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.
This paper expands on those identification results by presenting a new set of principal effect estimators, and associated standard error estimates, based on a general estimating equations (or M-estimation or method-of-moments) \citep{stefanskiBoos}.
It turns out that, under certain general conditions, principal effects can be estimated with a simple, two-step procedure: first, estimate principal scores with logistic regression or another M-estimator; then, impute unknown principal subgroup membership indicators with principal scores and estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We will refer to the estimators presented here as ``GEEPERs'': generalized estimating equations for principal effects using regressions.
GEEPERs estimates rely on correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, GEEPERs does not assume principal ignorability, and admits to sandwich-style standard error estimates, which we provide in the appendix.

@@

\section{Background}
In a randomized experiment with two conditions, let $Z_i=0$ if subject $i$, $i=1,\dots,n$ is assigned to the treatment condition, and let $Z_i=0$ otherwise.
If interest is in the effect of $Z$ on an outcome $Y$, under the ``stable unit treatment value assumption'' \citep{air} of no interference between units and no hidden versions of the treatment, define potential outcomes \citep{splawa1990application} $\yti$ as the value of $Y$ that $i$ would exhibit if $Z_i=1$, and let $\yci$ be the value if $Z_i=0$. Then the observed value of $Y$ satisfies $Y_i=Z_i\yti+(1-Z_i)\yci$, and the treatment effect of $Z$ on $Y$ for subject $i$ may be defined as $\tau_i=\yti-\yci$. The average treatment effect (ATE) is $\EE[\tau]$. %

@@

\subsection{Estimating Principal Effects}
In estimating principal effects for a binary $S$, it will be useful to define four conditional means. For $s\in\{0,1\}$, let 
\begin{equation}\label{eq:mus}
\mucs=\EE[Y_C|\st=s] \mbox{ and }\muts=\EE[Y_T|\st=s]%\mbox{ for }s=0,1
  % \begin{split}
  %   \muc0&=\EE[Y_C|\st=0]\\
  %   \muc1&=\EE[Y_C|\st=1]\\
  %   \mut0&=\EE[Y_T|\st=0]\\
  %   \mut1&=\EE[Y_T|\st=1]
  % \end{split}
\end{equation}
Then, since
\begin{equation*}
  \effs=\muts-\mucs
  % \begin{split}
  %   \eff0&=\mut0-\muc0\\
  %   \eff1&=\mut1-\muc1
  % \end{split}
\end{equation*}
estimating principal effects requires estimating the four conditional means.

@@

Under Assumptions \ref{ass:sm} and \ref{ass:rand}, conditional means $\mut0$ and $\mut1$ are non-parametrically identified.
In particular, the means of observed outcomes for subjects with $Z=1$ and $S=0$ or $S=1$ are unbiased for $\mut0$ and $\mut1$, respectively, since when $Z_i=1$, $Y_i=\yti$ and $S_i=\sti$.
In contrast, $\muc0$ and $\muc1$ are not fully identified without further assumptions because $\sti$ and $\yci$ are never observed simultaneously (though they may be nonparametrically bounded; see, e.g., \citealt{bounding}).
% When $S$ is a measure of compliance, and $\tau_i=0$ whenever $\sti=0$ (the ``exclusion restriction''), $\muc1$---and hence $\eff1$---may be estimated with instrumental variables methods \citep{air}.
% In this case, $\eff1$ is often called the ``local average treatment effect'' or the average effect of the treatment on the treated.
We will briefly review two approaches to estimating $\muc0$ and $\muc1$% that do not assume the exclusion restriction
---normal mixture modeling and weighting---with some comments in-between on the role for covariates.

\subsubsection{Normal Mixture Modeling}
The normal mixture modeling approach to estimating $\muc0$ and $\muc1$ \citep{imbens1997bayesian} assumes that, conditional on $\st$, $Y_C$ is normally distributed.
Then the probability density of $Y_C$ may be written as
\begin{equation}\label{eq:mixtureModel}
  f_{Y_C}(y)=Pr(\st=0)\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)+Pr(\st=1)\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
\end{equation}
where $\phi(\cdot)$ is the standard normal density function, and $\sigma_C^0>0$ and $\sigma_C^1>0$ are standard deviations, which are sometimes assumed equal.
The probabilities $Pr(\st=0)$ and $Pr(\st=1)$ may be estimated first using data from the treatment group, since Assumption \ref{ass:rand} implies that $Pr(\st=1|Z=1)=Pr(\st=1)$.
Given these estimates and \eqref{eq:mixtureModel}, parameters $\muc0$ and $\muc1$ may be estimated using maximum likelihood, method of moments, or Bayesian techniques.

@@

\subsection{Principal Scores}
The principal score \citep[e.g.][]{jo2009use} is defined as
\begin{equation}\label{eq:pscore}
  \ppi=Pr(\sti=1|\bxsi)
\end{equation}
where $\bxsi\subseteq\bm{x}_i$, the set of covariates used to model $S$.
Note that under randomization, $Pr(\sti=1|\bxsi,Z_i)=Pr(\sti=1|\bxsi)$, so a model for principal scores can be estimated using data from the treatment group and extrapolated to the control group.
Assume a model for principal scores:
\begin{equation}\label{eq:prinScoreMod}
  \ppip{\bm{\alpha}}=f(\bxsi;\bm{\alpha})
\end{equation}
with parameter vector $\bm{\alpha}$. An analyst may estimate $\bm{\alpha}$ by fitting model \eqref{eq:prinScoreMod} using observed $\bm{S}$ and $\bxs$ for subjects with $Z=1$, and then compute estimated principal scores $\ppip{\bm{\hat{\alpha}}}$ for subjects with $Z_i=1$.

@@

In order for principal scores to be useful, they need to vary with $\bm{x}$. Otherwise, $\bm{x}$ is uninformative about $\st$. In particular, we will assume the following:
\begin{ass}[Variable Principal Scores]\label{ass:vps}
 $\pp$ takes at least 3 distinct values
\end{ass}
%In some estimators we will discuss below, two distinct values suffice. 
For a broader discussion, see \citet{ding2011,jiangDing2021}.

Principal scores can potentially improve inference in a finite mixture model, by re-writing \eqref{eq:mixtureModel} as
\begin{equation}\label{eq:mixtureModelPS}
  f_{Y_C}(y)=(1-\ppip{\bm{\alpha}})\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)+\ppip{\bm{\alpha}}\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
\end{equation}
%Ideally, a statistician will estimate parameters $\muc0$, $\muc1$, $\mut0$, $\mut1$, $\sigma^0_C$, $\sigma^1_C$, and $\bm{\alpha}$ simultaneously, using maximum likelihood or Bayesian techniques.

Alternatively, an analyst may assume ``principal ignorability'' \citep{jo2009use,dingLu}:
\begin{ass}[Principal Ignorability]\label{ass:PI}
\begin{equation}\label{eq:pi}
  Y_{C}\independent \st |\bxs
\end{equation}
\end{ass}
or a somewhat weaker version, $\EE[Y_C|\st,\bxs]=\EE[Y_C|\bxs]$ \citep{feller2017principal}.
That is, the principal stratum is unrelated to control potential outcomes conditional on covariates.
Principal ignorability \eqref{eq:pi} is reminiscent of ignorability assumptions typical in observational studies \citep[e.g.][]{rosenbaum2010design}.
% Both require a rich set of covariates and background knowledge of the determinants of $S$; 
In particular, an unobserved covariate $u$ that is correlated with both $\st$ and $Y_C$ would invalidate \eqref{eq:pi}.

@@

We begin by introducing an assumption that we term ``Covariate Ignorability,'' or CI. It is a slightly weaker version of \citet{jiangDing2021}'s ``auxiliary independence'' assumption.
\begin{ass}[Covariate Ignorability]\label{ass:ci}
\begin{equation}\label{eq:assumption}
\EE[\yci|\bxsi,\sti]=\EE[\yci|\sti]=\muc0\text{ or }\muc1
\end{equation}
\end{ass}
i.e. $Y_C$ is mean-independent of $\bxs$ conditional on $\st$.
Under CI, covariates are not informative of the mean of $Y_C$ within principal strata.
As stated above, CI will rarely be plausible.
However, in some circumstances researchers can identify a subset of observed covariates that satisfy CI, and use those covariates in estimation.

It turns out that under strong monotonicity, randomization, and CI, and given a set of principal scores, the two principal effects $\eff0$ and $\eff1$ can be estimated by a simple OLS regression.
To see how, we'll %first
derive the estimating equations for $\muc0$, $\muc1$, $\mut0$, and $\mut1$, and show that they are equivalent to estimating equations for a model fit by OLS. % assuming that principal scores are known exactly, and then generalize to the case in which they are estimated using data from the treatment group.
The details of the calculations and the proofs are in the appendix.

The argument stems from a set of expressions for expectations, summarized in the following lemma:
\begin{lemma}\label{lemma:expectation}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, and \ref{ass:ci}, %if principal scores $\ppi$, for $i:Z_i=0$ are known exactly, then:
  \begin{equation*}
    \begin{split}
      \EE[Y|Z=0]&=\muc0+\EE\pp(\muc1-\muc0)\\
      \EE[Y\pp|Z=0]&=\EE[\pp]\muc0+\EE[\pp^2](\muc1-\muc0)\\
      \EE[Y|Z=1]&=\mut0+\EE[\st](\mut1-\mut0)\\
      \EE[YS|Z=1]&=\EE[\st]\mut0+\EE[\st^2](\mut1-\mut0)
    \end{split}
  \end{equation*}
\end{lemma}
Since $\st=\st^2$, the expression for $\EE[\st Y_T]$ in Lemma \ref{lemma:expectation} can be simplified to $\EE[\st]\mut1$; however, we retain the unsimplified version, since its similarity with the analogous expression when $Z_i=0$ simplifies the reasoning below.

@@

\begin{prop}\label{prop:reg1}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, \ref{ass:ci}, and suitible regularity conditions \citep[e.g.][p. 327]{boosStefanskiBook}, if %principal scores are known exactly,
  $\ri$ is defined as in \eqref{eq:ri} and the model
\begin{equation}\label{eq:regression0}
  \EE[Y_i|Z_i,\ri]=\beta_0+\beta_1\ri+\beta_2Z_i+\beta_3Z_i\ri
\end{equation}
is fit with OLS, yielding coefficient estimates $\hat{\beta}_k$, $k=0,\dots,3$, then
\begin{equation}
  %\begin{split}
    \heff0\equiv \hat{\beta}_2\mbox{ and }
    \heff1\equiv \hat{\beta}_3-\hat{\beta}_2
%  \end{split}
\end{equation}
are M-estimators, with $\heff0\rightarrow_p\eff0$ and $\heff1\rightarrow_p\eff1$ as $n\rightarrow\infty$
\end{prop}

@@

% If principal scores were known exactly, the M-estimation structure of $\heff0$ and $\heff1$ implies that their sampling variance could be estimated by the standard sandwich estimator for OLS \citep{stefanskiBoos}.
% This will rarely be the case.
%However, 
If the principal scores are modeled as \eqref{eq:prinScoreMod}, and parameters $\bm{\alpha}$ are estimated by M-estimation, with estimating equations $\sum_{i: Z_i=1}\Omega_i=\bm{0}$, then the estimating equations for the principal score model and principal effect estimation can be stacked \citep[c.f.][]{boosStefanskiBook}, as
\begin{equation}\label{eq:stacked}
  \sum_{i=1}^n \begin{pmatrix}
    Z_i\Omega_i(\bxsi,S_i;\bm{\alpha})\\
    \Psi_i\left(Y_i,Z_i,\ri(\bm{\alpha}); \bm{\beta}\right)\end{pmatrix}=\begin{pmatrix} \bm{0}\\\bm{0}\end{pmatrix}
\end{equation}
where $\ri\left(\bm{\alpha}\right)=S_iZ_i+\ppip{\bm{\alpha}}(1-Z_i)$ emphasizes $\ppi$'s dependence on $\bm{\alpha}$.
In practice, the principal score model \eqref{eq:prinScoreMod} could be fit first, yielding estimates $\bm{\hat{\alpha}}$, and model \eqref{eq:regression0} could be fit next, using $\ri=\ri\left(\bm{\hat{\alpha}}\right)$.
Together, the vector of estimates $[\bm{\hat{\alpha}},\bm{\hat{\beta}}]$ represents a zero of the stacked estimating equations \eqref{eq:stacked}.

Then, the standard error matrix for $[\bm{\hat{\alpha}},\bm{\hat{\beta}}]$ can be estimated as \citep[][ch. 7]{boosStefanskiBook}:
\begin{equation}\label{eq:sandwich}
  \widehat{var}(\bm{\hat{\beta}})=A^{-1}BA'
\end{equation}
Where
\begin{equation*}
  A=\sum_i\frac{\partial}{\partial [\alpha,\beta]'} [Z_i\Omega_i,\Psi_i]'
\end{equation*}
and
\begin{equation*}%\label{eq:Bmat}
  B=\sum_i [Z_i\Omega_i, \Psi_i]'[Z_i\Omega_i, \Psi_i]
\end{equation*}
where $[Z_i \Omega_i, \Psi_i]$ is the stacked vector of estimating equations evaluated at $Z_i$, $\bxsi$, $S_i$, and $Y_i$.
The appendix gives details of this calculation, giving a consistent variance estimator.

To summarize, under strong monotonicity, randomization, and covariate ignorability \eqref{eq:assumption}, principal effects can be estimated with the following procedure:
\begin{enumerate}
\item Estimate parameters $\bm{\alpha}$ as $\bm{\hat{\alpha}}$ by fitting model \eqref{eq:prinScoreMod} using data from subjects with $Z=1$.
\item Estimate principal scores for subjects with $Z_i=0$ as $\ppip{\bm{\hat{\alpha}}}$
\item Set $\ri=Z_iS_i+(1-Z_i)\ppip{\bm{\hat{\alpha}}}$
\item Using OLS, fit model \eqref{eq:regression0}
\item Estimate $\heff0=\hat{\beta}_2$ and $\heff1=\hat{\beta}_3-\hat{\beta}_2$ with sampling variance \eqref{eq:sandwich}
\end{enumerate}

\subsection{Relaxing Covariate Ignorability with Outcome Modeling}\label{sec:regression}
In most applications, identifying a set of covariates $\bxs$ satisfying CI will be difficult or impossible.
Fortunately, CI may be relaxed or obviated by regression (see \citealt[][\S 3.4]{jiangDing2021} for an analogous identification result).
Let $\bxy\in \mathbb{R}^k$, be a vector of covariates (possibly the same as $\bxs$); $\bxy$ can contain transformations and interactions of baseline covariates, allowing for a wide range of models.
By randomization, (Assumption \ref{ass:rand}), $\EE[\bxy|\st]=\EE[\bxy|\st,Z=0]=\EE[\bxy|\st,Z=1]$, so 
estimates of principal-stratum-specific means of $\bxy$ can be extrapolated from the treatment to the control group.
This leads to the following proposition:
\begin{prop}\label{prop:interactions}
  Say, for $i=1,\dots,n$, principal scores $\ppi$ are generated as \eqref{eq:pscore}, with parameters $\bm{\alpha}$ identified and estimable with M-estimation, and outcomes are generated as
  \begin{equation}\label{eq:interaction}
    \begin{split}
    Y_i=&\beta_0+\beta_1\sti+\beta_2 Z_i+\beta_3Z_i\sti\\
    &+\bm{\gamma_1}'\bxy_i+\bm{\gamma_2}'\bxy_i\sti+
    \bm{\gamma_3}'\bxy Z_i+\bm{\gamma_4}'\bxy_i Z_i\sti+\epsilon_i
    \end{split}
  \end{equation}
  with $\EE[\epsilon_i|\bxi,Z_i]=\EE[\epsilon_i|\bxi,Z_i,\sti]=0$ and $\epsilon_i\independent\epsilon_j$ when $i\ne j$.
